Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow

### 2014-08-14 writeups/dissertation/dissertation.tex
foo

So it may be worthwhile to use the standard form instead of the optimized learning form when declarative memory is modeled across a broader range of tasks.
However, this should not be taken as a strong recommendation,
since the performance benefits seen in this study could simply be due to the fact that the declarative memory retrievals span across a large period of time, which is not the case for all tasks.
Nonetheless, the only real reason not to use the standard form is due to computational resource limitations.
Since this is becoming less of an issue as time progresses and hardware speed increases, it may be time to reconsider the default form of base-level learning used in ACT-R.
It may be the case that the optimized learning form of the equation is no longer computationally necessary.
And if the standard form is used as the default (or simply tried in another experiment),
the results from this study suggest that the optimal decay rate value should be set slightly higher (\num{0.8}) than the default when optimized learning is used (\num{0.5}).

@@

Model accuracy was tested on the popular-hashtags dataset for Twitter and a random sample of posts across the entire StackOverflow dataset.
For each of the four Twitter popular-hashtags datasets, model retrievals were done across 15 runs and 500 posts within each run.
For the StackOverflow dataset, 5 runs of 100 posts were used, since the results for StackOverflow are more stable and retrievals take more wall-clock time for this dataset due to the size of the co-occurrence matrix.
All runs contained different posts, and all posts used for runs were not used to generate the co-occurrence matrix for the two retrieval models.

@@

The impact of adding the offset to each of the models for StackOverflow and Twitter are included in Figures \ref{figContextOffsetSO} and \ref{figContextOffsetT}.
The error bars represent the 95\% bootstrapped confidence interval of the mean model accuracy across all runs in dataset.

@@

In both plots, if the model name includes ``full'' then all model components are included.
When they are enumerated, then at least one of the components is left out.
The random permutation models are represented as ``RP'', Bayesian models as ``Bayes'', and they are plotted alongside each other so that comparisons across the two model types can be done easily.
So, for example, the ``standard prior model relaxed across posts without offset'' means that only the prior model component was used, and the offset for that component was not subtracted off when determining activation.
As another example, the ``Bayes only orderless context'' means that only the context component of the Bayesian model was used (along with the offset), and that no order information was used for that contextual component term.

Looking across the plots, including the offset significantly increases model accuracy, for all models, terms, and sites, except that the full Bayesian model without an offset was slightly more accurate on Twitter
(\num{.53} to \num{.60} and \num{.22} to \num{.45} for Bayesian and random permutation for StackOverflow, \num{.31} to \num{.29} and \num{.24} to \num{.25} for Bayesian and random permutation for Twitter).
Accuracy is improved even when only a single term is used.
Although in this case the weight of the term is irrelevant (only one term, weighting for that term does not matter),
using an offset still improves accuracy since the offset term ensures that only a small number of tags per posts are predicted by the model to be the chosen tags.

@@

The four subsets were collected across a period of one month.
The hashtags used in each subset were chosen by scraping a site that published the current trending hashtags for each city.
Hashtags for each US city were collected, and then 400 were randomly sampled from this set.
Only 400 were selected since this is the maximum that the Twitter API can monitor at one time.

@@

\textcite{Sahlgren2008} looked at several ways to handle the commonly-occurring stop words for the random permutation model:
A data-driven frequency-weighted approach was tried, along with using a previously-derived stoplist, and no stop word removal at all.
Removing the high-frequency words in the dataset showed the highest performance increase compared to no stop word removal, and using a previously-derived stoplist hardly increased performance.
However, \citeauthor{Sahlgren2008} did not try to weight the stop words, and only tried methods to remove them.
So alongside looking at frequency weighting, previously derived, and no stop word removal, \textcite{Stanley2013}'s method of attenuating stop words based on entropy was also explored.

@@

The original plot shows the size of the spikes when high-frequency words are not removed.
The entropy plot shows how the size decreases by more than an order of magnitude when the entropy weighting measure is used.
The cutoff for the frequency plot was chosen such that the plot looks similar to the entropy plot, with spikes around the same size.
Note that using a predefined stoplist does not remove some of the remaining high-frequency words in these datasets, which hints that this method may not be ideal when model accuracy is examined.
The cutoff that produced the frequency plot in Figure \ref{figContextCutoffT} is where a word represents more than \num{.04}\% of total occurrences in the dataset.
This may or may not be equal to the value for \textcite{Sahlgren2008}, since that number was expressed in total observations (not as a percentage), and the total number of words in the dataset was not given.

@@

Comparing across Figures \ref{figContextStopRPSO} and \ref{figContextStopRPT}, entropy and frequency techniques are superior to filtering based on a predetermined list.
However, it is more difficult to choose the best approach when comparing entropy weighting and frequency filtering.
Frequency filtering is slightly better for StackOverflow, while entropy weighting is slightly better for Twitter.
However, frequency filtering does have a few drawbacks compared to entropy weighting:
It produces accuracy measurements that have higher levels of noise (see Figure \ref{figContextStopRPSO} and the frequency model for the title component in StackOverflow).
Also, the ideal cutoff threshold must be identified in order to filter based on frequency,
and identifying this cutoff requires a search, and this search must be done for each dataset that frequency filtering is used on.
This cutoff value is a free parameter in frequency filtering that is not required when weighting stop words by entropy.
Therefore, favoring parsimony, for the random permutation model, attenuating stop words based on entropy produces more accurate results when compared to filtering the stop words based on frequency of use.
This result is interesting since filtering stop words based on frequency is the more common default method for random permutation models \parencite{Sahlgren2008}.
This suggests that a simpler method (entropy weighting) can be used instead of the more common frequency-filtering technique to achieve the same (or better) accuracy measurements.

@@

Another interesting result was how the Bayesian model actually performed worse when stop words were removed based on frequency.
This was not the case for the random permutation model in this study, was not the case when \textcite{Sahlgren2008} used frequency filtering on the random permutation model,
and is most likely never the case for the random permutation model.
Model accuracy most likely decreases for the Bayesian model because this model already has a normalization component built in that the random permutation model does not have.
When computing activation for the random permutation model, the correlation is calculated directly on a matrix of counts of co-occurrences for words and tags.
For the Bayesian model this count matrix is converted to a log-odds ($S_{ji}$) matrix,
where both the number of observations for the word and the tag are normalized when computing the value for each cell (see the $S_{ji}$ computation in Table \ref{tabModACTRModel}).
So it is likely the case that a form of frequency attenuation is already built into the Bayesian model.
And therefore adding a frequency filter on top of this does nothing to improve accuracy, and only removes predictive information.

@@

Another parameter in the random permutation model is the number of rows used to represent the all of the different words.
Each word is represented by two randomly placed ones and two negative ones across the rows.
All that is required is that the four positions and signs of each of these environment vectors is unique.
But this means multiple environment vectors can overlap when looking at a single row.
This is a lossy form of compression, and allows the number of rows to represent the words to be much less than the total number of words.
As the number of rows decreases, the amount of cross talk between the environment vectors increases, compression increases, and the amount of information stored decreases.

@@

In the figure, the full random permutation model that does not have the number of rows in the matrix specified is the \num{2048}-row matrix.
This was done so that names in all of the plots would remain consistent across all plots, and the \num{2048}-row matrix is the size used for the random permutation model for all other results.
The plot is also broken out for different size datasets used to build the co-occurrence matrix.

@@

The activation value for the context component of the vector-based model is a correlation, ranging from zero to one.
This is not a log-odds value.
For the previous results, prior and context for the random permutation model are combined by simply adding the terms.
However, this means that ACT-R's prior base-level learning component (which is a log-odds value) is added to a correlation value, which may not be mathematically appropriate.
It would be more appropriate to convert the correlation value to a log-odds value, and then add this context term to the log-odds prior term.

@@

The original BEAGLE model \parencite{Jones2007} actually showed only minimal improvement when incorporating word order as well.
In absolute terms, this model only improved by 2 percentage points in accuracy when trained on the TASA and tested on the TOEFL.
\textcite{Sahlgren2008}'s results for word order for the random permutation model are similar.
At a row size of around \num{2000}, model accuracy improves by less than a percentage point when order information is incorporated.
The results for larger row sizes are somewhat inconsistent, since the full model (order and orderless) sometimes underperforms the order-only model, which does not make sense.
\textcite{Sahlgren2008} most likely did not use a regression technique to optimize the weights for each of the order and orderless model terms.
So although incorporating word order into the random permutation model does seem to slightly improve performance,
the strength of the model is more likely to be in how efficiently it can represent the co-occurrence matrix, than its ability to include word order in the representation.

@@

Separate co-occurrence matrices were built for each user in the sampled StackOverflow and Twitter popular-users dataset.
The same sampling method used to test the impact of the prior component compared to the context component for the popular-users datasets was used,
but the oldest posts that were not included in the testing set were used to build the co-occurrence matrices.
So for each user, the co-occurrence matrix was created from their oldest posts, up to (and not including) their 50 newer posts that were used to test model accuracy.

@@

The declarative memory model for ACT-R has an attentional weight term $W_{j}$ that can be used to implement how stop words are handled.
The random permutation model was also easily modified to have an attentional weight term by attenuating each word's set of ones and negative ones values by that word's attentional weight.
Both a method for stop-word removal (frequency filtering) and stop-word attenuating (entropy weighting) performed well for both models.
Also, using a proper method for handling the low-predictor words produced some of the largest improvements in model accuracy compared to the other explorations
(e.g., compared to word order, properly adding terms, compression amount for random permutation representation).
Since the ACT-R declarative memory theory already has a component for attentional weight, it may be the case that the neural system performs some form of stop-word weighting when computing activations,
and this is reflected in the attentional weight term.
That is, $W_{j}$ can be defined mathematically, has the same general form across tasks, and both the frequency-filtering and entropy-weighting methods for stop-word removal are likely first candidates for the definition.
This claim, however, would require that no top-down activation is used to compute the attentional weight, which would mean that the $W_{j}$ is not under conscious control, which is not likely to be correct.
Perhaps it is a combination of top-down and bottom-up systems that determine $W_{j}$, and the entropy-weighting and frequency-filtering techniques are methods of formalizing the bottom-up component.

>>>>

So it may be worthwhile to use the standard form instead of the optimized learning form when declarative memory is modeled across a broader range of tasks.
However, this should not be taken as a strong recommendation,
since the performance benefits seen in this study could simply be due to the fact that the declarative memory retrievals span across a large period of time, which is not the case for all tasks.
Nonetheless, the only real reason not to use the standard form is due to computational resource limitations.
Since this is becoming less of an issue as hardware speed increases, it may be time to reconsider the default form of base-level learning used in ACT-R.
It may be the case that the optimized learning form of the equation is no longer computationally necessary.
And if the standard form is used as the default (or simply tried in another experiment),
the results from this study suggest that the optimal decay rate value should be set slightly higher (\num{0.8}) than the default when optimized learning is used (\num{0.5}).

@@

Model accuracy was tested on the popular-hashtags dataset for Twitter and a random sample of posts across the entire StackOverflow dataset.
For each of the four Twitter popular-hashtags datasets, model retrievals were done across 15 runs and 500 posts within each run.
For the StackOverflow randomly-sampled dataset, 5 runs of 100 posts were used,
since the results for StackOverflow are more stable and retrievals take more wall-clock time for this dataset due to the size of the co-occurrence matrix.
All runs contained different posts, and all posts used for runs were not used to generate the co-occurrence matrix for the two retrieval models.

@@

The impact of adding the offset to each of the models for StackOverflow and Twitter is shown in Figures \ref{figContextOffsetSO} and \ref{figContextOffsetT}.
The error bars represent the 95\% bootstrapped confidence interval of the mean model accuracy across all runs in dataset.

@@

In both plots, if the model name includes ``full'' then all model components are included.
When they are enumerated, then at least one of the components is left out.
The random permutation models are represented as ``RP'', Bayesian models as ``Bayes'', and they are plotted alongside each other so that comparisons across the two model types can be done easily.
So, for example, the ``standard prior model relaxed across posts without offset'' means that only the prior model component was used and the offset for that component was not subtracted off when determining activation.
As another example, the ``Bayes only orderless context'' means that only the context component of the Bayesian model was used (along with the offset), and that no order information was used for that contextual component term.

Looking across the plots, including the offset significantly increases model accuracy, for all models, terms, and sites
(\num{.53} to \num{.60} and \num{.22} to \num{.45} for Bayesian and random permutation for StackOverflow, \num{.31} to \num{.29} and \num{.24} to \num{.25} for Bayesian and random permutation for Twitter),
except that the full Bayesian model without an offset was slightly more accurate on Twitter.
Accuracy is improved even when only a single term is used.
Although in this case the weight of the term is irrelevant (only one term, weighting for that term does not matter),
using an offset still improves accuracy since the offset term ensures that only a small number of tags per posts are predicted by the model to be the chosen tags.

@@

The four subsets were collected across a period of one month.
Each was collected sequentially, and about one subset was collected per week.
The hashtags used in each subset were chosen by scraping a site that published the current trending hashtags for each city.
Before starting to collect for each subset, the current trending popular hashtags for each US city were collected, and then 400 were randomly sampled from this set.
Only 400 were selected since this is the maximum that the Twitter API can monitor at one time.

@@

\textcite{Sahlgren2008} looked at several ways to handle the commonly-occurring stop words for the random permutation model:
A data-driven frequency-weighted approach was tried, along with using a previously-derived stoplist, and no stop word removal at all.
Removing the high-frequency words in the dataset showed the highest performance increase compared to no stop word removal, and using a previously-derived stoplist hardly increased performance.
However, \citeauthor{Sahlgren2008} did not try to weight the stop words and only tried methods to remove them.
So alongside looking at frequency weighting, previously derived, and no stop word removal, \textcite{Stanley2013}'s method of attenuating stop words based on entropy was also explored.

@@

The original plot shows the size of the spikes when high-frequency words are not removed.
The entropy plot shows how the size decreases by more than an order of magnitude when the entropy weighting measure is used.
The cutoff for the frequency plot was chosen such that the plot looks similar to the entropy plot, with spikes around the same size.
Note that using a predefined stoplist does not remove some of the remaining high-frequency words in these datasets, which hints that this method may not be ideal when model accuracy is examined.
The cutoff that produced the frequency plots in Figures \ref{figContextCutoffSO} and \ref{figContextCutoffT} is where a word represents more than \num{.04}\% of total occurrences in the dataset.
This may or may not be equal to the value for \textcite{Sahlgren2008}, since that number was expressed in total observations (not as a percentage), and the total number of words in the dataset was not given.

@@

Comparing across Figures \ref{figContextStopRPSO} and \ref{figContextStopRPT}, entropy and frequency techniques are superior to filtering based on a predetermined list.
However, it is more difficult to choose the best approach when comparing entropy weighting and frequency filtering.
Frequency filtering is slightly better for StackOverflow, while entropy weighting is slightly better for Twitter.
However, frequency filtering does have a few drawbacks compared to entropy weighting:
It produces accuracy measurements that have higher levels of noise (see Figure \ref{figContextStopRPSO} and the frequency model for the title component in StackOverflow).
Also, the ideal cutoff threshold must be identified in order to filter based on frequency,
and identifying this cutoff requires a search, and this search must be done for each dataset that frequency filtering is used on.
This cutoff value is a free parameter in frequency filtering that is not required when weighting stop words by entropy.
Therefore, favoring parsimony, for the random permutation model, attenuating stop words based on entropy is a slightly superior method compared to removing stop words based on frequency of use.
This result is interesting since filtering stop words based on frequency is the more common default method for random permutation models \parencite{Sahlgren2008}.
This suggests that a simpler method (entropy weighting) can be used instead of the more common frequency-filtering technique to achieve the same (or better) accuracy measurements.

@@

Another interesting result was how the Bayesian model actually performed worse when stop words were removed based on frequency compared to no stop word removal.
This was not the case for the random permutation model in this study, was not the case when \textcite{Sahlgren2008} used frequency filtering on the random permutation model,
and is most likely never the case for the random permutation model.
Model accuracy most likely decreases for the Bayesian model because this model already has a normalization component built in that the random permutation model does not have.
When computing activation for the random permutation model, the correlation is calculated directly on a matrix of counts of co-occurrences for words and tags.
For the Bayesian model this count matrix is converted to a log-odds ($S_{ji}$) matrix,
where both the number of observations for the word and the tag are normalized when computing the value for each cell (see the $S_{ji}$ computation in Table \ref{tabModACTRModel}).
So it is likely the case that a form of frequency attenuation is already built into the Bayesian model.
And therefore adding a frequency filter on top of this does nothing to improve accuracy, and only removes predictive information.

@@

Another parameter in the random permutation model is the number of rows used to represent the all of the different words.
Each word's environment vector is represented by two randomly placed ones and two negative ones across the rows.
All that is required is that the four positions and signs of each of these environment vectors is unique.
But this means multiple environment vectors can overlap when looking at a single row.
This is a lossy form of compression, and allows the number of rows to represent the words to be much less than the total number of words.
As the number of rows decreases, the amount of cross talk between the environment vectors increases, compression increases, and the amount of information stored decreases.

@@

In the figure, the model name that does not have the number of rows in the matrix specified is the \num{2048}-row matrix.
This was done so that names in all of the plots would remain consistent across all plots, and the \num{2048}-row matrix is the size used for the random permutation model for all other results.
The plot is also broken out for different size datasets used to build the co-occurrence matrix.

@@

The activation value for the context component of the vector-based model is a correlation, ranging from zero to one.
This is not a log-odds value.
For the previous results, prior and context for the random permutation model were combined by simply adding the terms.
However, this means that ACT-R's prior base-level learning component (which is a log-odds value) is added to a correlation value, which may not be mathematically appropriate.
It would be more appropriate to convert the correlation value to a log-odds value, and then add this context term to the log-odds prior term.

@@

The original BEAGLE model \parencite{Jones2007} actually showed only minimal improvement when incorporating word order as well.
In absolute terms, this model only improved by 2 percentage points in accuracy when trained on the TASA and tested on the TOEFL.
\textcite{Sahlgren2008}'s results for word order for the random permutation model are similar.
At a row size of around \num{2000}, model accuracy improves by less than a percentage point when order information is incorporated.
The results for larger row sizes are somewhat inconsistent, since the full model (order and orderless) sometimes underperforms the order-only model, which does not make sense.
\textcite{Sahlgren2008} most likely did not use a regression technique to optimize the weights for each of the order and orderless model terms.
So although incorporating word order into the random permutation model does seem to slightly improve performance,
the strength of the model is more likely to be in how efficiently it can represent the co-occurrence matrix rather than its ability to include word order in the representation.

@@

Separate co-occurrence matrices were built for each user in the sampled StackOverflow and Twitter popular-users dataset.
The same sampling method used to test the impact of the prior component compared to the context component for the popular-users datasets was used,
and the co-occurrence matrices were built from the oldest posts that were not included in the testing set.
So for each user, the co-occurrence matrix was created from their oldest posts, up to (and not including) their 50 newer posts that were used to test model accuracy.

@@

The Bayesian declarative memory model for ACT-R has an attentional weight term $W_{j}$ that can be used to implement how stop words are handled.
The random permutation model was also easily modified to have an attentional weight term by attenuating each word's set of ones and negative ones values by that word's attentional weight.
Both a method for stop-word removal (frequency filtering) and stop-word attenuating (entropy weighting) performed well for both models.
Also, using a proper method for handling the low-predictor words produced some of the largest improvements in model accuracy compared to the other explorations
(e.g., compared to word order, properly adding terms, compression amount for random permutation representation).
Since the ACT-R declarative memory theory already has a component for attentional weight, it may be the case that the neural system performs some form of stop-word weighting when computing activations,
and this is reflected in the attentional weight term.
That is, $W_{j}$ can be defined mathematically, has the same general form across tasks, and both the frequency-filtering and entropy-weighting methods for stop-word removal are likely first candidates for the definition.
This claim, however, would require that no top-down activation is used to compute the attentional weight, which would mean that the $W_{j}$ is not under conscious control, which is not likely to be correct.
Perhaps it is a combination of top-down and bottom-up systems that determine $W_{j}$, and the entropy-weighting and frequency-filtering techniques are methods of formalizing the bottom-up component.
